## Effects of Severe Weather on Property and Population in the US for the Time Period 1950 to November 2011

blueFedora

8/24/2014

### Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This analysis involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The data were downloaded and subsetted to contain only the pertinent variables.  Some minor cleanup was performed and the data then summarized and the top 20 event types for damage and casualties are listed and the top 10 event types are graphed for your convenience.

### Data Processing
#### Setup the needed libraries and environment.

```{r setEnvironment}
library(utils)
library(plyr)
library(ggplot2)
library(gridExtra)
opts_chunk$set(echo=TRUE, fig.width=11, fig.height=6)
events <- 20 # number of event types for which to report results
```

#### Download data and documentation files

```{r 'downloadData'}
Download <- function (fileURL, destfile, method="auto", quiet=TRUE) {
    if(!file.exists("./data")) {
        dir.create("./data")
    }
    destfile <- paste0("./data/", destfile)
    if (!file.exists(destfile)) {
        if (grepl("https:", fileURL) & method=="auto") {
            method <- "curl"
        }
        print(sprintf("DL file: %s on %s", destfile, date()))
        download.file(fileURL, destfile=destfile, method=method, quiet=quiet)
    }
}
#download data file
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
dest <- "StormData.bz2"
Download(url, dest)

#download storm data documentation
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
dest <- "pd0101600tcurr.pdf"
Download(url, dest)

#download storm events FAQ
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf"
dest <- "StormEventsFAQ.pdf"
Download(url, dest)
```

#### Read data into StormData data.frame

```{r 'readData', cache=TRUE}
StormData <- read.csv("./data/StormData.bz2")
```

Exploring the storm data:

```{r 'exploreStormData'}
dim(StormData)
names(StormData)
```

#### Build a working data.frame and minor data clean up

We'll build a working data.frame and include only the pertinent columns, EVTYPE, CROPDMG, CROPDMGEXP, PROPDMG, PROPDMGEXP, FATALITIES and INJURIES. Also, there are "Summary" EVTYPEs.  These are not "storms" as such and will be removed from the working data. 

```{r subsetData}
workData <- subset(StormData, !grepl("Summary", StormData$EVTYPE), select=c('EVTYPE', 'CROPDMG', 'CROPDMGEXP', 'PROPDMG', 'PROPDMGEXP', 'FATALITIES', 'INJURIES'))
```

#### More clean up and damage calculations

Doing a summary of the exponent data for crops and property damage, CROPDMGEXP and PROPMGEXP variables, we get:

```{r summaryData}
summary(workData$CROPDMGEXP)
summary(workData$PROPDMGEXP)
```

The NOAA documentation states that the exponent values of the crop and property damage (CROPDMGEXP and PROPDMGEXP) are alpha and are: "K" for thousands, "M" for millions, "B" for billions, other values such as numbers and symbols are not defined.

Characters such as the sequence 0 through 8, "H", "+", "-", "?" and lower case m, k, h are not mentioned in the documentation.  This analysis in general takes a conservative approach toward the data, does NOT make assumptions and will ignore the observations for all of these undocumented exponent characters.

New fields for the crop and property damage exponents, CEXP and PEXP, will be set to the proper power of ten based on the valid exponent value.  An invalid exponent value will result in a value of NA which utlimately results in that observation being excluded from the results.

```{r 'calculateEventTypeTotals'}
#first fill the EXP variables with NAs
workData$CEXP <- NA
workData$PEXP <- NA

#then set the exponents to the proper power of ten based on the letters K, M, B

workData$CEXP <- ifelse(grepl('K', workData$CROPDMGEXP), 10^3, workData$CEXP)
workData$CEXP <- ifelse(grepl('M', workData$CROPDMGEXP), 10^6, workData$CEXP)
workData$CEXP <- ifelse(grepl('B', workData$CROPDMGEXP), 10^9, workData$CEXP)

workData$PEXP <- ifelse(grepl('K', workData$PROPDMGEXP), 10^3, workData$PEXP)
workData$PEXP <- ifelse(grepl('M', workData$PROPDMGEXP), 10^6, workData$PEXP)
workData$PEXP <- ifelse(grepl('B', workData$PROPDMGEXP), 10^9, workData$PEXP)

#calculate the damages to crops and property per event type observation
workData$CROPDMG <- workData$CROPDMG * workData$CEXP
workData$PROPDMG <- workData$PROPDMG * workData$PEXP

#calculate totals by event type
totals <- ddply(workData, ~ EVTYPE, summarize,
                CropDmgImpact=sum(CROPDMG, na.rm=TRUE),
                PropDmgImpact=sum(PROPDMG, na.rm=TRUE),
                TotalDmgImpact=sum(CROPDMG, PROPDMG, na.rm=TRUE),
                Fatalities=sum(FATALITIES, na.rm=TRUE),
                Injuries=sum(INJURIES, na.rm=TRUE),
                TotalHealthImpact=sum(FATALITIES, INJURIES, na.rm=TRUE))
```

### Results

#### Health Impact Reports

Top `r events` events for fatalities, injuries and total health impact (fatalities + injuries) per event type.

```{r healthReports}
print(totals[order(-totals$Fatalities),c(1,5)][1:events, ], row.names=FALSE)
print(totals[order(-totals$Injuries),c(1,6)][1:events, ], row.names=FALSE)
print(totals[order(-totals$TotalHealthImpact),c(1,7)][1:events, ], row.names=FALSE)
```

```{r healthGraph}
#plotData <- data.frame()
#plotData[, 1] <- totals[order(-totals$Fatalities),c(1)][1:5]
#plotData[, 2] <- totals[order(-totals$Fatalities),c(5)][1:5]
#qplot(plotData[, 1], plotData[, 2], geom='bar', stat='identity')
plot1 <- 
    qplot(totals[order(-totals$Fatalities),c(1)] [1:5],
      totals[order(-totals$Fatalities),c(5)] [1:5]/1000, 
      main='Fatalities',
      geom='bar', 
      xlab="Event Type",
      ylab="Thousands of Fatalities",
      stat='identity')
plot1 <- plot1 + theme(axis.text.x = element_text(angle=90))

plot2 <- 
    qplot(totals[order(-totals$Injuries),c(1)] [1:5],
      totals[order(-totals$Injuries),c(6)] [1:5]/1000, 
      main='Injuries',
      geom='bar', 
      xlab="Event Type",
      ylab="Thousands of Injuries",
      stat='identity')
plot2 <- plot2 + theme(axis.text.x = element_text(angle=90))

plot3 <- 
    qplot(totals[order(-totals$TotalHealthImpact),c(1)] [1:5],
      totals[order(-totals$TotalHealthImpact),c(7)] [1:5]/1000, 
      main="Total Health Impact",
      geom='bar', 
      xlab="Event Type",
      ylab='Thousands of (Injuries + Fatalities)',
      stat='identity')
plot3 <- plot3 + theme(axis.text.x = element_text(angle=90))


#x <- totals[order(-totals$Fatalities),c(1)][1:5]

#y <- x[1:5]
#y
#x
#d <- totals[order(-totals$Fatalities),c(1)][1:5]
#str(d)
#d
grid.arrange(plot1, plot2, plot3, ncol=3)
```
We see from the above printed tables and graphs that the top threats for human fatalities are TORNADO, EXCESSIVE HEAT and FLASH FLOOD.  The top threats for human injuries are TORNADO, TSTM WIND and FLOOD.  The top combined (fatalities and injuries) health threats come from TORNADO, EXCESSIVE HEAT and TSTM WIND.

We notice that there are event types which are similar; e.g., Hurricane and Hurricane/Typhoon; Excessive Heat and Heat; and Flood, River Flood and Flash Flood.  Although it may seem logical, we have no evidence that these event types are actually the same and made the decision to not combine them. 

#### Damage Impact Reports
Top `r events` events for crop damage, property damage and total damage (crop damage + property damage)
```{r damageReports}
print(totals[order(-totals$CropDmgImpact), c(1,2)][1:events, ], row.names=FALSE)
print(totals[order(-totals$PropDmgImpact), c(1,3)][1:events, ], row.names=FALSE)
print(totals[order(-totals$TotalDmgImpact),c(1,4)][1:events, ], row.names=FALSE)
```

```{r damageGraph}
```

We see from the above printed tables and graphs that the top threats for crop damage are DROUGHT, FLOOD and RIVER FLOOD.  The top threats for property damage are FLOOD, HURRICANE/TYPHOON and TORNADO.  The top combined (crop and property damage) threats come from FLOOD, HURRICANE/TYPHOON and TORNADO.

### Computing Environment
```{r 'sessionInfo'}
sessionInfo()
```